---
title: "Bycatch impacts exploration tool: report"
output: html_document
params:
  lh.params: NA
  performance: NA
  pbr: NA
---

This report was generated using the [marine mammal bycatch impacts exploration tool](https://msiple.shinyapps.io/mammaltool/). The output reflects parameters that the user has entered for their marine mammal population of interest. These values are entered in the  **Advanced** and **PBR & PBR calculator** tabs. 

### Life history parameters used

```{r echo=FALSE,  warning=FALSE, message=FALSE}
# The `params` object is available in the document.
options(knitr.table.format = "html") 
library(magrittr)

lhdf <- as.data.frame(params$lh.params)

lhdf <- lhdf %>%
  dplyr::select(S0,S1plus,AgeMat,lambdaMax,z) %>%
  dplyr::rename("$S_0$" = S0,
         "$S_{1+}$" = S1plus,
         "Age at maturity" = AgeMat,
         "$\\lambda_{max}$" = lambdaMax)

knitr::kable(format="html", # make table showing all the parameters
             lhdf) %>% 
            kableExtra::kable_styling()
```

### Performance measures
These are the median outcomes for each of the bycatch and depletion levels. 

```{r echo=FALSE,  warning=FALSE, message=FALSE}
# The `params` object is available in the document.
pm <-  params$performance 

knitr::kable(format="html",pm) %>%
  kableExtra::kable_styling(full_width = F,position="center")

```

### Estimated reference points

The following table is based on values that the user has entered in the **PBR** tab of the app. 

If $N_{MIN}$ and $N_{BEST}$ are the same, the user has not specified a survey CV of abundance. If several values are missing, the user may have run some projections without entering anything in the **PBR** tab. 

**Note:** For this table to give the correct values, the user has to specify a current population size and an annual bycatch rate in the **Advanced** tab. Otherwise, the numbers below will not be based on a real population size. 


```{r echo=FALSE,  warning=FALSE, message=FALSE}
# The `params` object is available in the document.
PBR <- params$pbr %>%
  dplyr::mutate(Parameter = dplyr::recode(Parameter, 
                            Nbest = "$N_{BEST}$",
                            Nmin = "$N_{MIN}$",
                            Rmax = "$R_{MAX}$",
                            Fr = "$F_R$"))
PBR %>% 
  dplyr::mutate_if(is.numeric, list(~as.character(., signif(., 2)))) %>%
  knitr::kable(format="html") %>%
  kableExtra::kable_styling(full_width = F,position="center")

```